---
title: "Bfiat: Modelling trawling effects on ecosystem functioning"
author: "Karline Soetaert and Olivier Beauchard"
output:
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
    code_folding: show
    theme: cosmo
abstract: "Bfiat is an R-package to estimate the impact of bottom trawling on organism density at the taxon level, and on ecosystem functioning. In this vignette we demonstrate the use of the package using simple examples."
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Bfiat: Modelling trawling effects on ecosystem functioning}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Bfiat)
palette("ggplot2")  # ggplot2-style palette
Colramp <- ramp.col(c("lightgreen", "darkblue"), n=100) # colors for image
```

# Aim. 

This document describes how to estimate the impact of bottom fisheries on sediment ecosystem functioning, using the R-package *Bfiat* (Bottom Fisheries Impact Assessment Tool). 

This R-package contains a data-driven mechanistic model that describes how species density is being reduced by bottom trawling, and then recovers till the next fishing event. 

The outcome of this biological module are trajectories of species (taxon) densities over time. As the species densities change, also the ecosystem functions delivered by the community change.  


# Two logistic growth models including fisheries 

Logistic growth describes how species density or biomass evolves when the sediment is undisturbed (i.e. in between fishing events). 
The logistic model reads:

$$\frac{dD_i}{dt}=r_i \cdot D_i \cdot (1-\frac{D_i}{K_i}),$$

where $t$ is time, $D_i$ is the density of species $i$ at a particular time, $K_i$ is the carrying capacity of species $i$, $r_i$ is the logistic growth parameter (units [/time]). 
The term on the left hand side, $\frac{dD_i}{dt}$ expresses how $D_i$ changes over time (it is a derivative).

The parameters $r_i$ and $K_i$ are specific for species $i$; the carrying capacity $K_i$ also depends on the site where the species is found. 

## Fisheries events

To include the impact of fisheries in the logistic growth model, we can assume that trawling occurs as a series of *events*, that instantaneously decimate populations. In between these events the populations recover. 
Such fishing models consist of two parts:

1. When the sediment is trawled, at times $t_j$, the density of each species $i$, is instantaneously depleted with a fraction $d_i$, so that only a fraction $1-d_i$ remains:

$$D_i^{t_j^+} = D_i^{t_j^-} \times (1-d_i),$$
Here $D_i^{t_j^+}$ and $D_i^{t_j^-}$ are the density after and before trawling occurs at time $t_j$ respectively, and $d_i$ is the depletion fraction. 
The higher is the depletion factor $d_i$, the more the density of the species after trawling will have been reduced.

2. In between fishing, the population restores according to the logistic growth equation 
$$\frac{dD_i}{dt}=r_i \cdot D_i \cdot (1-\frac{D_i}{K_i}).$$

The \emph{annual swept area ratio, S}, the proportion of a grid cell swept by fishing gear per year, is used to estimate the mean *time in between fishing events* : $\Delta T = \frac{1}{S}$, in years. From this, the times of the fishing events can be estimated, e.g. as $T_j = T_{j-1} + \Delta T$, where we assume equally spaced fishing events. 

## Continuous fishing

Another approoach to model fishing pressure on populations is to assume that fishing induces a constant mortality on the benthos of a large area rather than being caused by a set of localised events. 

In this case, the (continuous) dynamic model reads (e.g. Pitcher et al., 2017):
$$\frac{dD_i}{dt}=r_i \cdot D_i \cdot (1-\frac{D_i}{K_i}) - m_i  \cdot D_i,$$

where $m_i$ is the constant mortality induced upon the species $i$.

# R-package Bfiat 

The R-package *Bfiat* (Soetaert and Beauchard, 2024) contains:

* Functions to estimate bottom trawling impacts on density 
```{r, echo = FALSE}
ll <- ls("package:Bfiat")
lf <- ll[grep("_", ll)]

l1 <- lf[!grepl("get_", lf)]
l1[!grepl("par_", l1)]
```

* Functions to estimate bottom trawling impacts on ecosystem functions.
```{r, echo = FALSE}
lf[grepl("get_", lf)]
```
* Functions to calculate fishing parameters.
```{r, echo = FALSE}
lf[grepl("par_", lf)]
```
* Datasets to apply the Bfiat functions.
```{r, echo=FALSE}
ll[!grepl("_", ll)]
```

```{r, echo=FALSE}
Def_fun <- function(name){
  
  zn <- ifelse(is.function(name), substitute(name), name)
  Z <- try(deparse(args(name)), silent=TRUE)
  if (inherits(Z, "try-error")) {
    cat(paste(zn, ": dataset\n\n"))
  } else {

  Z[1] <- substr(Z[1], 9, nchar(Z[1]))
  Z[1] <- paste(zn,Z[1])
  Zl <- Z[length(Z)]
  Z[length(Z)] <- substr(Zl, 9, nchar(Zl)-4)

  for (i in 1:10)  #  remove double blancs
    Z <- gsub("  ", " ", Z)
  
  for (i in 1:length(Z))
    cat(Z[i], "\n")
  }
}
```

# Two functions for the assessment of bottom trawling on benthic density. 

## Trawling events

Functions *density_perturb* and *run_perturb* estimate the impact of bottom disturbances on benthic taxa. Both functions implement the same model, but require slightly different inputs. For instance, function *run_perturb* accepts randomly spaced trawling times (via argument *events*), while *density_perturb* does not.

Function *density_metier*  allows to input several metiers, each with their own sar and depletion fraction.

They are defined as:

```{r, echo=FALSE, comment=NA}
Def_fun(density_perturb)
Def_fun(density_metier)
Def_fun(run_perturb)
```

As an example, we estimate trawling impact on one species, with trawling occurring either once or twice per year ($S1=1, S2=2$). The trawling lasts for 10 years, after which the sediment remains untrawled for another 10 years.

```{r, fig.cap="Two trawling impact runs, for a species with [r=0.5/yr, K=10 ind, d=0.1]; trawling lasts for 10 years, followed by a recovery of 10 years"}
times <- seq(0, 20, length.out=1000)   # output times, years

# species-specific parameters
parms <- list(r = 0.5,   # [/yr], rate of increase
              K = 10,    # [ind], carrying capacity
              d = 0.1)   # [-]  , depletion fraction

# run the model - one trawling event per year
out1 <- density_perturb(parms          = parms,
                        times          = times, 
                        tstart_perturb = 2,
                        tend_perturb   = 10,
                        sar            = 1.)  
               
# run the model - two trawling events per year
out2 <- density_perturb(parms          = parms,
                        times          = times, 
                        tstart_perturb = 2,
                        tend_perturb   = 10,
                        sar            = 2.)  
# show the results
plot(out1, out2, 
     xlab = "years", 
     ylab = "density", 
     main = "Trawling with events", 
     lty = 1, lwd = 2, las=1)

legend("bottomright", 
       title = "sar",legend = c("1/yr", "2/yr"), 
       lty = 1, col = 1:2)
```

## Constant trawling

Functions *density_logistic* and *run_logistic* implement a fishing model with constant mortality, and are defined as:

```{r, echo=FALSE, comment=NA}
Def_fun(density_logistic)
Def_fun(run_logistic)
```

We run this model with mortality = 0.1 and 0.2 /year. 
As before we assume that fishing stops after 10 years. This can simply be set with argument *tend_perturb*.

```{r, fig.cap="Two trawling impact runs, for a species with [r=0.5/yr, K=10 ind, d=0.1, m=0.1 and m=0.2/yr]; trawling lasts for 10 years, followed by a recovery of 10 years"}

times <- seq(0, 20, length.out=1000)   # output times, years

parms <- list(r = 0.5,   # [/yr], rate of increase
              K = 10,    # [ind], carrying capacity
              m = 0.1)   # [/yr], mortality rate

c1 <- density_logistic(times          = times, 
                       tstart_perturb = 2,
                       tend_perturb   = 10, 
                       parms          = parms)

# increase mortality
p2      <- parms
p2$m    <- 0.2

c2 <- density_logistic(times          = times, 
                       tstart_perturb = 2,
                       tend_perturb   = 10, 
                       parms          = p2)

# Show results
plot(c1, c2, 
     xlab = "years", ylab = "density", 
     main = "Constant trawling", 
     lty = 1, lwd = 2, las = 1)

legend("bottomright", 
       legend = c("0.2/yr", "0.4/yr"), title = "m",
       lty = 1, col = 1:2)
```

## Effect of trawling on several species

The same models can be run for several species at once:

```{r, fig.width=6, fig.height=10, fig.cap="Trawling impacts on three species (s1, s2 and s3) and for two trawling intensities"}
times <- seq(0, 20, length.out=1000)  # output times      [year]

# Species names and parameters
parms <- data.frame(
  taxon = c("s1", "s2", "s3"),
  K     = c(   2,   10,    4),  # carrying capacity [g/m2]
  r     = c(   2,    5,  0.4),  # rate of increase  [/year]
  d     = c( 0.5,  0.3,  0.1)   # depletion rate
)

# run the model
run1 <- density_perturb(parms          = parms, 
                        times          = times, 
                        tstart_perturb = 2,
                        tend_perturb   = 10,
                        sar            = 0   )

run2 <- density_perturb(parms          = parms, 
                        times          = times, 
                        tstart_perturb = 2,
                        tend_perturb   = 10,
                        sar            = 1   )

run3 <- density_perturb(parms          = parms, 
                        times          = times, 
                        tstart_perturb = 2,
                        tend_perturb   = 10,
                        sar            = 2   )
# show the results
plot(run1, run2, run3, 
     lty=1, lwd=2, mfrow=c(3,1), las=1)
legend(x="bottomright", 
       legend=c("0", "1", "2"), title="trawls/year", 
       col=1:3, lty=1, lwd=2)
```


## Multiple metiers

We can also simulate several metiers (fishing activities), each with their own depletion induced on the species:

```{r, fig.width=10, fig.height=10, fig.cap="Trawling impacts on four species (s1, s2, s3 and s4) and for three metiers"}
times <- seq(0, 20, length.out=1000)  # output times      [year]

# Species names and parameters
parms <- data.frame(
  taxon = c("s1", "s2", "s3", "s4"),
  K     = c(   2,  10,    4,     1),  # carrying capacity [/m2]
  r     = c(   5,   2,  0.4,     1)   # rate of increase  [/year]
)

# 4 metiers

sar  <- data.frame(m1 = 1, m2 = 0.7, m3 = 1.5)

# depletion rate is species (rows) and metier- specific (columns)
#                          sp1   sp2    sp3    sp4
d    <- data.frame(m1 = c( 0.5,  0.3,   0.1,   0.3),  # metier 1
                   m2 = c( 0.3,  0.6,   0.2,   0.2),  # metier 2
                   m3 = c( 0.2,  0.1,  0.02,  0.05))
# run the model
runm <- density_metier(parms          = parms, 
                       d              = d,
                       times          = times, 
                       tstart_perturb = c(0.5, 1, 1.5),
                       tend_perturb   = 10,
                       sar            = sar   )
# show the results
plot(runm, lty = 1, las = 1) #, main = "3 metiers")
```

# Trawling impacts on traits and ecosystem functions

## Trawling impacts on traits

Function *get_trait_model* combines modeled species responses with traits; it is defined as:

```{r, echo=FALSE, comment=NA}
Def_fun(get_trait_model)
```

Using the previous model runs, we estimate the impact of trawling on the feeding type of species, where we assume they are either suspension or deposit feeders.

We first define the extent to which, for each species, both types of feeding contributes to its diet, and where the sum of the proportions = 1.

```{r}

feedingtype <- data.frame(taxon      = c("s1", "s2", "s3"), 
                          suspension = c(   0,  0.5,   1), 
                          deposit    = c(   1,  0.5,   0))

knitr::kable(feedingtype)
```

Then we calculate how both feeding types are affected by the trawling. 
We first estimate the total numbers of both types, after which we also estimate the relative contribution (using argument *scalewithvalue*).

```{r, fig.width=8, fig.height=8, fig.cap="Change in absolute  (top) and relative (bottom) contribution of feeding types due to trawling"}
FT  <- get_trait_model(model          = run2, 
                       trait          = feedingtype, 
                       scalewithvalue = FALSE)

FT2 <- get_trait_model(model          = run2, 
                       trait          = feedingtype)

plot(FT, ylab = "/m2",  
     mfrow = c(2, 2), las = 1)
plot(FT2, ylab = "-",
     main = c("fraction suspension feeding", "fraction deposit feeding"), 
     mfrow = NULL, las = 1)
```

Similarly, we can merge the model output for instance with mean weight of each taxon, where when *scalewithvalue=FALSE*, we will obtain the summed weights of all individuals, whereas if *scalewithvalue=TRUE*, we will obtain the mean weight of the entire community.

```{r, fig.width=8, fig.height=4, fig.cap="Change in mean and total weight due to trawling"}
# Biomass as a trait
WW   <- data.frame(taxon   = c("s1", "s2", "s3"), 
                   biomass = c(0.1,   0.5,   1))
Lrun  <- get_trait_model(model = run2, 
                         trait = WW, 
                         scalewithvalue=TRUE)
Lrun2 <- get_trait_model(model = run2, 
                         trait = WW, 
                         scalewithvalue=FALSE)
plot(Lrun,  main = "mean weight", xlab = "g/ind",  
     mfrow = c(1, 2), las = 1)
plot(Lrun2, main = "summed weight", xlab = "g/m2",
     mfrow = NULL, las = 1)
```

## Trawling impacts on ecosystem functions

Functions *get_Db_model*,  *get_irr_model* estimate the effect of species responses on bioturbation and bioirrigation rates, as described in Querios et al., 2013 and Wrede et al., 2018 respectively. They are defined as:

```{r, echo=FALSE, comment=NA}
Def_fun(get_Db_model)
Def_fun(get_irr_model)
```

To calculate the *bioturbation index*, we need to input the relevant trait information. 
This should be a data.frame, with the first column the name of the taxon, and followed by columns denoting the mobility  and reworking modes (values between 1-4 and 1-5 respectively). These columns should be named "Mi" and "Ri". 

For the *bioirrigation index*, next to the taxon name the columns named "BT" (burrowtype, value from 1-3), "FT" (feeding type, 1-3), and "ID" (injection depth, 1-4) should be present. 

To calculate both indices, we also require the mean weight of the taxa; this should be passed as a data.frame with taxonname in the first column, and the weight in the second.

The formula for estimating the bioturbation Index for taxon i (as in Querios et al., 2013) is:
$$BPc_i = \sqrt(W_i)  \cdot  density_i  \cdot  R_i  \cdot  M_i$$

The formula for estimating the bioirrigation Index for taxon i (as in Wrede et al., 2018) is:
$$IPc_i = W_i^{0.75}  \cdot  density_i  \cdot  BT_i  \cdot FT_i \cdot ID_i$$,

As an example, we first input the weights and traits required, and then estimate the bioturbation and bioirrigation potential associated to previous runs.

The bioturbation and bioirrigation function estimate the contribution of each species to these indices, as well as the total index. 

```{r, fig.width=6, fig.height=6, fig.cap="Change in absolute  (top) and relative (bottom) contribution of feeding types due to trawling"}
weight   <- data.frame(taxon  = c("s1", "s2", "s3"), 
                       weight = c( 0.5,  1.0,   3))

Dbtrait  <- data.frame(taxon = c("s1", "s2", "s3"), 
                       Ri    = c(   4,    3,    1), 
                       Mi    = c(   3,    3,   2))

irrtrait <- data.frame(taxon = c("s1", "s2", "s3"), 
                       BT    = rep(1, times=3), 
                       FT    = c(   3,  2.5,   2), 
                       ID    = c(   3,    3,   1))

Db  <- get_Db_model(model  = run2, 
                    trait  = Dbtrait, 
                    weight = weight)
plot(Db, las=1)

Irr <- get_irr_model(model  = run2, 
                     trait  = irrtrait, 
                     weight = weight)
plot(Irr, las=1)
```

# Functions to estimate fishing parameters. 

The package *Bfiat* also contains functions to estimate the fishing parameters.

## Depletion rate

Function *par_d* estimates the depletion rate as a function of the depth occurrence of a species and of the gear penetration depth. 

The *depletion fraction* ($d_i$) for a certain species $i$ is implemented as:
$$d_i = min \left[m_{max}, \sum_{z=[z_u, z_l]}  m_d \cdot max\left( 0, p_G-z_u\right) \cdot fi_{[z_u, z_l]}\right]$$
where $z_u$ and $z_l$ are the upper and lower depth of the species' living space, $m$ is a coefficient that scales the depletion to the depth that the gear has penetrated in the species's living space (the sediment depth inbetween $z_u, z_l$), and where $fi_{[z_u, z_l]}$ is the fractional occurrence of species $i$ in that zone. The maximal depletion fraction is limited to the value $m_{max}$. 

Good values for these parameters are $m_d = 0.055$ $cm^{-1}$ and $m_{max} = 0.42$. The function is defined as:

```{r, echo=FALSE, comment=NA}
Def_fun(par_d)
```

Below, the depletion rate is estimated for three species, where the living depth of the species has been divided in 2 sediment layers (0-10 and 10-20 cm), and is fuzzy coded.

The first species lives in the upper sediment layer (D0_10) only; the second is equally distributed between layers and the third is only found below 10 cm:
 
```{r}
 fDepth <- data.frame(D0_10  = c(1, 0.5, 0),   #upper depth=0
                      D10_20 = c(0, 0.5, 1))   #upper depth=10
 uDepth <- c(0, 10)  # upper depth                           
``` 

The depletion, using the default parameter settings, and for a gear penetration depth of 1 cm is:

```{r}
 par_d (gpd    = 1, 
        fDepth = fDepth, 
        uDepth = uDepth)

 par_d (gpd    = 10, 
        fDepth = fDepth, 
        uDepth = uDepth)

```
Note that the deep-living species is not affected by any of the gears. 

## Rate of increase

Function *par_r* estimates the rate of increase (units $time^{-1}$), as a function of the species age at maturity (a_i) as follows: $$r_i=\frac{4.165}{a_i}.$$
It is defined as:

```{r, echo=FALSE, comment=NA}
Def_fun(par_r)
```

```{r, fig.width=6, fig.height=4, fig.cap="relationship between longevity and rate of increase"}
 Age_mat <- seq(from = 0, to = 5, length.out=100)
 plot(x    = Age_mat, 
      y    = par_r(Age_mat), 
      xlab = "age at maturity, year", 
      ylab = "r, /year", 
      main = "relationship between age at maturity and r",
      las  = 1)
```

## Carrying capacity

Function *par_K* estimates the species carrying capacity, based on the fishing parameters, and species density. 
It does so by inverse modelling, assuming that the measured density is the density of the fauna, under the fishing pressure (sar).

```{r, echo=FALSE, comment=NA}
Def_fun(par_K)
```

```{r}
 par_K (density =   1, 
        sar     =   1, 
        r       = 0.1, 
        d       = c(0.05, 0.3))
```

Not all combinations of parameters lead to viable carrying capacities, notably if the mortality induced by fisheries is larger than the species intrinsic rate of increase (*r*), then the species should become extinct due to fishing.

Carrying capacity K can be estimated for a sequence of fishing intensities

```{r, fig.width=6, fig.height=4, fig.cap="relationship between carrying capaicty and swept area ratio"}
# sequence of fishing intensities
 sar.seq <- seq(from=0.05, to=3, by=0.05)

# estimate carrying capacity for all these fishing intensities
 K.seq   <- par_K(density = 1,
                  sar     = sar.seq,
                  r       = 0.5,
                  d       = 0.2)
 
 plot(sar.seq, K.seq,      
      xlab = "Swept area ratio, /yr", 
      ylab = "K/D",
      main = "relationship between sar and K",
      las = 1, ylim = c(0, 10))
```

Note that the higher the trawling intensity (or swept area ratio), the more the current density will deviate from the actual carrying capacity, so the K/sar ratio significantly increases with sar. Above a sar=2.25, the mortality becomes larger than the rate of increase, and the species is likely to be extinct.

## Continuous mortality coefficient

Function *par_m* estimates the constant mortality parameter for a species with growth rate *r*, so that the continuous fishing model is consistent with the discrete model with parameters *d* and the fishing intensity *sar*.

```{r, echo=FALSE, comment=NA}
Def_fun(par_m)
```

```{r}
par_m(sar = 2, 
      r   = 0.5, 
      d   = 0.2, K = 1)
```

# Accessory functions

## Density at pertubation, and steady-state density

Function *density_perturb* estimates densities at selected times, while *trawl_perturb* estimates the density before and after the n-th trawling event.

```{r, echo=FALSE, comment=NA}
Def_fun(density_perturb)
Def_fun(trawl_perturb)
```

```{r}
density_perturb(sar = 1, 
                r   = 0.5, 
                d   = 0.2, 
                times  = seq(0, 2, by = 0.5))

trawl_perturb(sar = 2, 
              r   = 0.5, 
              d   = 0.2, 
              n   = 10:12)

trawl_perturb(sar = 2:10, 
              r   = 0.5, 
              d   = 0.2, 
              n   = 1)

```

Function *steady_perturb* and *steady_logistic* estimate the density at steady state.

```{r, echo=FALSE, comment=NA}
Def_fun(steady_perturb)
Def_fun(steady_logistic)
```

```{r}
(sD <- steady_perturb(K   = 1,
                      sar = c(1, 2, 3),
                      r   = 0.5, 
                      d   = 0.2))

# this is the inverse and should return K=1
par_K(density = sD$density,
      sar     = c(1, 2, 3),
      r       = 0.5, 
      d       = 0.2)
```

# Datasets

There are 2 data sets in the *Bfiat* package:

## Density and biomass data

Dataset *NIOZdata* provides macrofauna density and biomass data at 4 sites in the NorthSea.

```{r, fig.width=6, fig.height=6, fig.cap="Stations in the NIOZdata dataset"}
map_key(x        = NIOZdata$stations$x,
        y        = NIOZdata$stations$y, 
        contours = NIOZdata$contours, colkey = FALSE,
        col.levels = "grey", by.levels = 5, draw.levels = TRUE, 
        pch = 16)

text(x      = NIOZdata$stations$x, 
     y      = NIOZdata$stations$y+0.5, 
     labels = NIOZdata$stations$station)
```

The characteristics of NIOZdata, in terms of density and biomass are:

```{r, echo=FALSE}
SD <- get_summary(data       = NIOZdata$density, 
                  descriptor = station,
                  taxon      = taxon, 
                  value      = density)
           
SD2 <- get_summary(data       = NIOZdata$density, 
                   descriptor = station,
                   taxon      = taxon, 
                   value      = biomass_ww)

TD <- merge(SD$density, SD2$density, by=1)
TD <- merge(TD, SD$taxa)
names(TD) <- c("station", "density", "biomass", "numTaxa")
TD <- merge(TD, NIOZdata$abiotics)
knitr::kable(TD, digits=c(rep(0, times=6), rep(2, times=4)), 
             caption="Characteristics of NIOZdata: density (ind/m2), biomass (gWW/m2), number of taxa, depth (m), % silt, sar (/yr), subsar (/yr) and gear penetration depth (cm)")
```

## Swept Area Ratio

Fishing intensity for the Northsea area as annual swept area ratios are in a dataset called *SAR*, with a resolution of 0.05 dg in longitudinal, and 0.05 dg in the latitudinal direction. 
The data originate from ICES, and have been compiled on request of OSPAR. 
They comprise:

* *sar*:  the swept area ratio of the surface 0-2 cm
* *subsar*: the swept area ratio of the subsurface (> 2cm),
* *gpd*, the mean gear penetration depth, estimated based on the different metiers, i.e. 4.45, 1.55, 1.55, 2.55 cm for dredge, otter trawl, seine, beam trawl respectively (the mean of the gear penetration in mud and sand).
* *lon*, *lat*, the coordinates of the points

```{r, fig.width=8, fig.height=12, echo = FALSE, caption="Fisheries intensity for the Northsea"}
par(mfrow=c(2,2), oma=c(0,0,2,0))

DD   <- NSBS$contours
keep <- which(DD$x < max(SAR$x) & 
              DD$x > min(SAR$x) & 
              DD$y < max(SAR$y) & 
              DD$y > min(SAR$y) & 
              DD$z==0 | is.na(DD$x))

with (SAR, image2D(
     x = x, y = y, z=sar, 
     main = "surface trawling [0-2 cm]", 
     clab = "#/yr", 
     xlab = expression(""^o~E), 
     ylab = expression(""^o~N),
     log = "c", asp = 1.7, las = 1, clim=c(0.0001, 50),
     colkey = list(length = 0.5, width = 0.5), 
     col    = Colramp
))
lines(DD$x[keep],DD$y[keep])

with (SAR, image2D(
      x = x, y = y, z = subsar, 
      main = "deep trawling [> 2 cm]",      
      clab = "#/yr", 
      xlab = expression(""^o~E), 
      ylab = expression(""^o~N),
      log = "c", asp = 1.7, las = 1, clim = c(0.0001, 50),
      colkey = list(length = 0.5, width = 0.5), 
      col    = Colramp
))
lines(DD$x[keep],DD$y[keep])

with (SAR, image2D(
      x = x, y = y, z = gpd, 
      main = "gear penetration depth", 
      clab = "cm",      
      xlab = expression(""^o~E), 
      ylab = expression(""^o~N),
      log = "c", asp = 1.7, las = 1, clim = c(1.5, 4.5),
      colkey = list(length = 0.5, width = 0.5), 
      col    = Colramp
))
lines(DD$x[keep],DD$y[keep])

CC <- alpha.col(4:5, alpha=0.5)
hist(x      = na.omit(log10(as.vector(SAR$sar))),
     breaks = 100, 
     main   = "frequency of SAR [0-2cm]", 
     xlab   = "SAR (/year))", 
     axes = FALSE, xlim = c(-3, 2), las = 1, col = CC[1], border = CC[1])

hist(x      = na.omit(log10(as.vector(SAR$subsar))), 
     breaks = 100,  
     col = CC[2], border = CC[2], add = TRUE)

axis(side   = 2)
axis(side   = 1, 
     at     = -3:2, 
     labels = 10^(-3:2))
legend("topleft", 
       legend = c("gpd 0-2cm", "gpd > 2cm"), 
       col = CC, pch = 16, pt.cex = 2)
mtext(outer = TRUE, side = 3,"Annual Swept Area Ratio")
```

Note: the R-package, *Btrait* (Soetaert and Beauchard, 2022) also contains density and trait data that can be used for fishery assessment. 

# Sensitivity and critical values

## Taxon sensitivity

Function *sensitivity_taxon* estimates for one taxon the density decrease for combinations of sar (swept area ratio) and gpd (gear penetration depth); also estimates (sar x gpd) values where the taxon reaches a critical level.

```{r}
Def_fun(sensitivity_taxon)
```

Function critical_sar and critical_gpd estimate the critical fishing intensity and gear penetration depth above which the taxon would become extinct (crit_DK = 0) or attain some predefined fraction of its carrying capacity (crit_DK > 0).

```{r}
Def_fun(critical_sar)
Def_fun(critical_gpd)
```

In the figure below, first the sensitivity for the species Arctica islandica in the Dutch North sea area (the MWTL data) is estimated and plotted. 
The sensitivity to fishing intensity (sar) and gear penetration depth (gpd) shows a rather large area where the species cannot survive (the white region). We then add all stations from the Dutch part of the North sea, characterised by the sar and gpd. This shows how the species would do in these stations. We end by estimating the critical sar for the gears (gpd) used in the Dutch Northsea, above which Artica islandica would dissappear.

```{r, fig.width=8, fig.height=6}
par(mfrow = c(1,1), las = 1)

# parameters for Arctica islandica
subset(MWTL$fishing, 
       subset = (taxon == "Arctica islandica"))

S_af <- sensitivity_taxon(r      = 0.64, 
                          fDepth = c(0.75, 0.25),
                          uDepth = c(0,      5))

# image, white = extinct
image2D(x = S_af$sar, y = S_af$gpd, z = S_af$DK, 
        xlab = "sar, /yr", ylab = "gpd, cm", 
        main = c("Species sensitivity", "Arctica islandica"),
        col = jet2.col(100), clab = "D/K")
lines(S_af$sar, S_af$critical_gpd, lwd = 2)
        
# add the stations in the MWTL data (from Btrait)
points(MWTL$abiotics$sar, MWTL$abiotics$gpd, 
       pch = 18, col = "black")
legend("topright", pch = 18, legend = " MWTL stations")

c_s_MWTL <- critical_sar(gpd = MWTL$abiotics$gpd, 
                         fDepth = c(0.75, 0.25),
                         uDepth = c(0,      5))
summary(c_s_MWTL)
```

## area sensitivity and critical values

Function *sensitivity_area* estimates for one set of sar and gpd values the density decrease for combinations of r (rate of natural increase) and d (depletion) of a taxon; also estimates critical (r x d) combinations for the area.

```{r}
Def_fun(sensitivity_area)
```

Functions critical_r and critical_d estimate the critical intrinsic rate of natural increase (r) and depletion fraction (d) beyond which a taxon, that experiences the fishing intensity and gear penetration depth of the area, would become extinct (crit_DK = 0) or attain some predefined fraction of its carrying capacity (crit_DK > 0).

```{r}
Def_fun(critical_r)
Def_fun(critical_d)
```

We plot the sensitivity for one station in the Dutch part of the Northsea, and add the species from the MWTL area on this plot.

```{r, fig.width=8, fig.height=6}
par(mfrow = c(1,1), las = 1)

# parameters for station WADDKT04
subset(MWTL$abiotics, 
       subset = (station== "WADDKT04"),
       select = c(station, sar, gpd))

S_W <- sensitivity_area(sar      = 11.817, 
                        gpd      = 1.90,
                        r.seq    = seq(0,  11, length.out = 200),
                        d.seq    = seq(0, 0.2, length.out = 200))


# image of station sensitivity, white = extinct (D/K = NA)
image2D(x = S_W$r, y = S_W$d, z=S_W$DK, 
        xlab = "r, /yr", ylab = "d, -", 
        main = c("Area sensitivity", 
                 "WADDKT04"),
        col = jet2.col(100), clab = "D/K")
lines(S_W$r, S_W$critical_d, lwd = 2)

# Species characteristics relating to fishgin for species in the MWTL data 
# (from package Btrait)
SP_Fish    <- MWTL$fishing

# depletion for all MWTL species in station WADDKT04
SP_Fish$dW <- par_d(
          gpd    = 1.90, 
          fDepth = SP_Fish[, c("p0", "p0_5cm", "p5_15cm", "p15_30cm", "p30cm")],
          uDepth =           c(  0,    0,        5,         15,         30))
points(SP_Fish$r, SP_Fish$dW, 
       pch = 18, col = "black")
legend("topright", pch = 18, legend = " MWTL species")
```

\newpage

# Fishing impact in a station in the North sea: FrieseFront

We now calculate the impact of bottom trawling on one station in the Norths Sea, located in the FrieseFront, and part of the NIOZdata dataset.

This station is located at `r subset(NIOZdata$stations, station=="FrieseFront")$x` dgE and `r subset(NIOZdata$stations, station=="FrieseFront")$y` dgN. 
It is `r subset(NIOZdata$abiotics, station=="FrieseFront")$depth` m deep, and has a fine-sandy sediment with median grain size = `r subset(NIOZdata$abiotics, station=="FrieseFront")$D50` $\mu m$. 

## Density and biomass

To apply the fisheries model to this station, we first select the *density* and *biomass* of the macrofauna. The biomasses are in wet weight and called "biomass_ww".

```{r}
FF <- 
  get_density(data       = NIOZdata$density,
              descriptor = station,
              subset     = station == "FrieseFront",
              taxon      = taxon,
              value      = data.frame(density, biomass_ww)
    )
```

## Fishing parameters

To estimate fishing effects, we need, for each taxon, the living depth (substratum depth distribution), and the age at maturity.

We first find the names, and meaning of the traits in the NIOZ trait database (Traits_nioz, from package *Btrait*), and select the traits relating to living depth and age at maturity:

```{r}
traits.lab      <- metadata(Traits_nioz)

# metadata of traits in database (name, value, ...)
meta.depth     <- subset(traits.lab, 
                         subset = trait == "Substratum depth distribution")

meta.maturity  <- subset(traits.lab, 
                         subset = trait == "Age at maturity")

# select actual traits
traits.depth    <- subset(Traits_nioz, 
                          select = c("taxon", meta.depth$colname))

traits.maturity <- subset(Traits_nioz, 
                          select = c("taxon", meta.maturity$colname))
```

We then extract the traits for each species in the data, passing also the taxonomic relationships. This allows the traits to be estimated also for species that are not documented in the trait database. The taxonomy has been obtained from the worms database, and is stored in *NIOZdata$taxonomy*.

The trait database records the traits in fuzzy format. For the depth distribution, we keep the data fuzzy, as some species may be found in different depth levels. 

As we need the *mean* age at maturity, we also pass the trait class and value (argument trait_class, trait_score).

```{r}
taxa <- unique(NIOZdata$density$taxon)

Depth    <- get_trait(taxon       = taxa, 
                      trait       = traits.depth, 
                      taxonomy    = NIOZdata$taxonomy)

colnames(Depth)[-1] <- meta.depth$modality

Maturity <- get_trait(taxon       = taxa, 
                      trait       = traits.maturity, 
                      taxonomy    = NIOZdata$taxonomy, 
                      trait_class = meta.maturity$trait,
                      trait_score = meta.maturity$value)

Traits_combined <- merge(Depth, Maturity)
```

The age at maturity can now be used to estimate the intrinsic rate of natural increase:

```{r}
Traits_combined$r <- par_r(age_at_maturity = Traits_combined$Age.at.maturity)
```


This information is merged with the data.frame that contains the station-species information, where the common columns are the "taxon":

```{r}
FF <- merge(FF, Traits_combined, 
            by = "taxon")
```

For each station in the data set, the *Swept Area Ratio* (sar) and estimated gear penetration depth (gpd), averaged over the years 2009-2017 is stored in the *NIOZdata$abiotics* table; the sar and gpd for station FrieseFront is extracted and added to the parameter file:

```{r}
FF$sar <- subset(NIOZdata$abiotics, 
                 subset = station=="FrieseFront")$sar
FF$gpd <- subset(NIOZdata$abiotics, 
                 subset = station=="FrieseFront")$gpd


FF <- FF[order(FF$density, decreasing=TRUE),]
row.names(FF) <- NULL
```

The *depletion* during trawling is estimated based on the depth of occurrence. 

The depletion rate can now be estimated, based on the sar (`r FF$sar[1]`) and the gear penetration depth (`r FF$gpd[1]`). We use the function *par_d* to estimate the depletion rate.

```{r}
fDepth  <- FF[,c("0", "0-5", "5-15", "15-30", ">30")]
uDepth  <-     c( 0,   0,     5,      15,      30)

FF$d <- par_d(gpd    = FF$gpd, # gear penetration depth
              fDepth = fDepth, # fractional occurrence in depth layers
              uDepth = uDepth) # upper depth of layers
```

We can now estimate the species carrying capacity, using function *par_K*.

```{r}
FF$K <- par_K(density = FF$density, 
              sar     = FF$sar, 
              r       = FF$r, 
              d       = FF$d)
```

The parameter file is now thinned, removing the columns we do not need any further:

```{r}
pnames <- c("taxon", "density", "biomass_ww", "K", "r", "d", "sar")
FF     <- FF[order(FF$taxon),pnames]
```

```{r}
knitr::kable(head(FF, n=6), digits=1, row.names = FALSE)
```

## Model run

The model is run for 40 years, where trawling occurs only the first 20 years. We specify the times at which a fishing event occurs and pass this to the model function (*run_perturb*). The argument *addsum*, when set equal to TRUE calculates the summed densities.

```{r}
# timings of the fishing events, based on sar
events <- seq(from = 1/FF$sar[1], 
              by   = 1/FF$sar[1], 
              to   = 20)

times  = seq(0, 40, length.out = 10000)

# run the model
fish.FF <- run_perturb(parms        = FF, 
                       times        = times, 
                       sar          = FF$sar[1],
                       tend_perturb = 20,
                       addsum       = TRUE)
```

As the parameter file was ordered with respect to the species density, it is easy to plot the density trajectory for the 8 most abundant species (*which=1:8*); this is followed by the total density.

```{r, fig.width=8, fig.height=8, fig.cap="Fishing impact on the density of the 8 most abundant species, and the total density in station FrieseFront; 20 years of fishing followed by 20 years without fishing."}

Abundant <- FF$taxon[order(FF$K, decreasing = TRUE)]

plot(fish.FF, 
     which = Abundant[1:8], 
     ylab  = "ind/m2", 
     xlab  = "year", las=1)

plot(fish.FF, 
     which = "sum", 
     ylab  = "ind/m2", 
     xlab  = "year", 
     las=1, mfrow=NULL)  # mforw=NULL to prevent it from starting a new figure
```

Using the previous model run, we estimate the impact of trawling on suspension feeders.

We first extract the modality "Suspension feeder" from the nioz trait database:

```{r}
meta.susp   <- subset(
                    metadata(Traits_nioz), 
                    subset= (modality == "Suspension feeder"))

traits.susp <- subset(Traits_nioz, 
                      select = c("taxon", meta.susp$colname))
colnames(traits.susp)[2] <- "SuspensionFeeder"
```

The nioz trait database stores the trait modalities in fuzzy format, so the data record the fractional contribution of suspension feeding to the diet of each species:

```{r}
head(traits.susp, n=3)
```

We now calculate the degree to which suspension feeding changes in the community, using the function *get_trait_model*.
For those species in the model whose suspension feeding is not known, we use the taxonomic closeness to find good values; this is simply done by passing data.frame *Taxonomy*.

```{r}
susp1 <- get_trait_model(model    = fish.FF, 
                         trait    = traits.susp, 
                         taxonomy = NIOZdata$taxonomy)
plot(susp1, 
     main = "Relative abundance suspension feeders", ylab = "-", las = 1)
```

The bioturbation potential can be easily estimated by passing the correct database to the function *get_Db_model*. For this function, we also need the mean weight of the species which is calculated first (*WW*).

```{r}
WW <- data.frame(taxon = FF$taxon, 
                 W     = FF$biomass_ww/FF$density)

Db.FF <- get_Db_model(model    = fish.FF, 
                      trait    = Traits_Db, 
                      taxonomy = NIOZdata$taxonomy, 
                      weight   = WW, 
                      na.rm    = TRUE)  # removes those species where W is NA
```

```{r, fig.width=8, fig.height=8, fig.cap="Fishing impact on the contribution to the bioturbation potential of the main species, and the total bioturbation potential in station FrieseFront; 20 years of fishing followed by 20 years without fishing."}
plot(Db.FF, 
     which = paste("BPc" , Abundant[1:8], sep="_"),
     ylab  = "BPc, -", las = 1)

plot(Db.FF, 
     which = "BPc", 
     ylab  = "-", 
     las = 1, mfrow = NULL)
```

# Acknowledgements

This product was created under the European Marine Observation Data Network
(EMODnet) Biology Phase V. Additional support came from the NWO-funded project BFIAT (NWO 18523).

# References

Beauchard O, Brind'Amour A, Schratzberger M, Laffargue P, Hintzen NT, Somerfield PJ, Piet G (2021) A generic approach to develop a trait-based indicator of trawling-induced disturbance. Mar Ecol Prog Ser 675:35-52. https://doi.org/10.3354/meps13840

Beauchard, O., Murray S.A. Thompson, Kari Elsa Ellingsen, Gerjan Piet, Pascal Laffargue, Karline Soetaert, 2023. Assessing sea floor functional diversity and vulnerability. Marine Ecology Progress Serie v708, p21-43, https://www.int-res.com/abstracts/meps/v708/p21-43/

Chamberlain S, Vanhoorne. B (2023). worrms: World Register of Marine Species (WoRMS) Client. R package
  version 0.4.3, https://CRAN.R-project.org/package=worrms.

Hiddink, JG, Jennings, S, Sciberras, M, et al. Assessing bottom trawling impacts based on the longevity of benthic invertebrates. J Appl Ecol. 2019; 56: 1075– 1084. https://doi.org/10.1111/1365-2664.13278 

Hiddink, Jan Geert, Simon Jennings, Marija Sciberras, Claire L. Szostek, Kathryn M. Hughes, Nick Ellis, Adriaan D. Rijnsdorp, Robert A. McConnaughey, Tessa Mazor, Ray Hilborn, Jeremy S. Collie, C. Roland Pitcher, Ricardo O. Amoroso, Ana M. Parma, Petri Suuronen, and Michel J. Kaiser, 2017.  Global analysis of depletion and recovery of seabed biota after bottom trawling disturbance. Proc. Nat. Aca. Sci, 114 (31) 8301-8306. https://doi.org/10.1073/pnas.161885811.

Queiros, A. M., S. N. R. Birchenough, J Bremner, J.A. Godbold, R.E. Parker, A. Romero-Ramirez, H. Reiss, M. Solan, P. J. Somerfield, C. Van Colen, G. Van Hoey, S. Widdicombe, 2013. A bioturbation classification of European marine infaunal invertebrates. Ecology and Evolution 3 (11), 3958-3985

Wrede, A., J.Beermann, J.Dannheim, L.Gutow, T.Brey, 2018. Organism functional traits and ecosystem supporting services - A novel approach to predict bioirrigation. Ecological indicators, 91, 737-743.

Soetaert K, Beauchard O (2024). R-package Btrait: Working with Biological density, taxonomy, and trait composition
data. Netherlands Institute of Sea Research. Data product created under the European Marine Observation Data Network
(EMODnet) Biology Phase IV.
  
Karline Soetaert, Olivier Beauchard (2024). Bfiat: Bottom Fishing Impact Assessment Tool. Data product created under the European Marine Observation Data Network (EMODnet) Biology Phase V.
  
R Core Team (2021). R: A language and environment for statistical computing. R Foundation for Statistical
  Computing, Vienna, Austria. URL https://www.R-project.org/.
  
C.R. Pitcher, J.G. Hiddink, S. Jennings, J. Collie, A.M. Parma, R. Amoroso, T. Mazor, M. Sciberras, R.A. McConnaughey, A.D. Rijnsdorp, M.J. Kaiser, P. Suuronen, and R. Hilborn, 2022.
Trawl impacts on the relative status of biotic communities of seabed sedimentary habitats in 24
regions worldwide. PNAS. https://doi.org/10.1073/pnas.2109449119

C.R. Pitcher, N. Ellis, S. Jennings, J.G. Hiddink, T. Mazor, M.J.Kaiser, M.I. Kangas, R.A. McConnaughey, A.M. Parma, A.D. Rijnsdorp, P. Suuronen, J.S. Collie, R. Amoroso, K.M. Hughes and R. Hilborn, 2017. Estimating the sustainability of towed fishing-gearimpacts on seabed habitats: a simple quantitative riskassessment method applicable to data-limited fisheries. Methods in Ecology and Evolution2017,8,472–480doi: 10.1111/2041-210X.12705

J. Depestele, K. Degrendele, M. Esmaeili, A. Ivanovic, S. Kroger, F.G. O’Neill, R. Parker, H. Polet, M. Roche, L.R. Teal, B. Vanelslander, A.D. Rijnsdorp. (2018). Comparison of mechanical disturbance in soft sediments due to tickler-chain SumWing trawl versus electro-fitted PulseWing trawl. ICES J. Mar. Sci.
76, 312–329.
  
# Appendices

## List of all functions and data sets in Bfiat:

```{r, echo=FALSE, comment=NA}
Bf <- ls("package:Bfiat")
for (i in 1:length(Bf))
  Def_fun(Bf[i])
```

## Estimating mean Life-span from fuzzy coded values

To calculate the rate of increase, we need the average life span of the organism. In many trait databases this characteristic is fuzzy coded, i.e. a taxon may belong to multiple modalities. 

This is the case for the nioz trait database (*Traits_nioz*).
To convert the data to crisp format, we first need to find the name of the trait that captures life span.

```{r}
NIOZ_att <- metadata(Traits_nioz)
unique(NIOZ_att$trait)

LSatt <- subset(NIOZ_att, subset=trait=="Life span")
LSatt
```

The column names of lifespan traits in the nioz database are in LS$colname. They are, together with the taxon name extracted from the database:

```{r}
LS <- Traits_nioz[, c("taxon", LSatt$colname)]
head(LS)
```


We now calculate the average life span using function *get_trait*. By passing the "trait_class" and "trait_score", the averages are estimated. We also pass the taxonomic tree of these data, so that the lifespan of taxa for which this is unknown is estimated based on taxonomic closeness.  

```{r}
MeanLS <- get_trait(taxon       = FF$taxon,
                    trait       = LS,
                    trait_class = LSatt$trait,
                    trait_score = LSatt$value,
                    taxonomy    = NIOZdata$taxonomy)
head(MeanLS)
```
