---
title: "The Bfiat package: modelling trawling effects in the Dutch part of the Northsea"
author: "Karline Soetaert and Olivier Beauchard"
output:
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
    code_folding: show
    theme: cosmo
abstract: "The impact of bottom trawling on benthic life in the Dutch part of the Northsea is estimated. We use monitoring data from 103 stations in the area (MWTL data set), and impose fishing intensities estimated for each of these stations."
vignette: >
 %\VignetteEncoding{UTF-8} 
 %\VignetteIndexEntry{Bfiat: The Bfiat package: modelling trawling effects in the Dutch part of the Northsea} 
 %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, knitr.table.format = 'rst')
library(Btrait)
library(Bfiat)
palette("ggplot2")                                      # ggplot2-style palette
Colramp <- ramp.col(c("lightgreen", "darkblue"), n=100) # colors for image
```

# Aim.

We use functions from the *Btrait* and *Bfiat* R-packages to estimate the impact of bottom fisheries on sediment ecosystem functioning in the Dutch part of the Northsea.

We use three sets of data:

*   *in situ* density data from a particular site.
*   species *trait* information, including:
    - the age at maturity of the species, and 
    - the depth of occurrence in the sediment.
*   fishing intensity and gear penetration depths in the sites

We use the MWTL macrofauna density data that have been compiled in the frame of the EMODnet biology project, (<https://www.emodnet-biology.eu>). These data are available from the R-package *Btrait* (Soetaert and Beauchard, 2024a).

The NIOZ trait database is used for the species traits (Beauchard et al., 2021, 2023).
The trait dataset is also in the R-package *Btrait*.

# Software.

The fishing impact models run in the open source framework *R* (R core team, 2025) and have been implemented in the *Bfiat* R-package (Soetaert and Beauchard, 2024b).

The R-package *Btrait* (Soetaert and Beauchard, 2022) contains functions to work on density and trait datasets.

The packages *Btrait* and *Bfiat* are available on github (<https://github.com/EMODnet/Btrait> and <https://github.com/EMODnet/Bfiat>).

# The fisheries impact model

The fisheries impact model is based on the logistic growth equation (Verhulst 1845) that describes how species density or biomass evolves over time. The effect of fishing on the population is included as a series of *events*, that instantaneously decimate population sizes, while in between these events the populations recover.

The model thus consists of two parts: (continuous) logistic growth in between fishing events and an instantaneous depletion during fishing.

## Logistic growth

The differential equation is, for t in between $t_j^+$ and $t_{j+1}^-$: 
$$\frac{dD_i^t}{dt}=r_i \cdot D_i^t \cdot (1-\frac{D_i^t}{K_i}),$$ 
where $t$ is time, $D_i^t$ is the density of species $i$ at a particular time, $K_i$ (units of density) is the carrying capacity of species $i$, $r_i$ is the logistic growth parameter (units [/time]).

## Instantaneous trawling

During trawling at times $t_j$, density changes instantaneously:

$$D_i^{t_j^+} = D_i^{t_j^-} \cdot (1-d_i),$$

where $D_i^{t_j^+}$ and $D_i^{t_j^-}$ are the density immediately after and before trawling occurs at time $t_j$ respectively, and $d_i$ is the species-specific and trawl-specific depletion fraction.

## Trawling intensity 

The time in between trawling can be inferred from the fishing intensity, expressed as the annual swept area ratio (SAR).

SAR is the cumulative area contacted by a fishing gear over one year and per surface area.
It has units of $m^2~m^{-2}~yr^{-1}$, or $yr^{-1}$.

From the swept area ratio (hereafter denoted as $S$), we calculate the average time in between fishing events ($yr$) as:

$$\Delta t = t_{j+1} - t_j = \frac{1}{S}.$$


\newpage

# Data requirements

Central in the impact analysis is the use of a combination of data sets:

## Community composition

Benthic biological data to be used for fishing assessment should comprise at least density data, and preferentially also biomass of individual species.

The densities or biomasses are used to estimate the "carrying capacity" of the species at a particular site; this is the abundance the species attains at steady-state, when undisturbed (i.e. in the absence of fishing).

## Taxonomy

To assign biological traits to the taxa in the community, it is convenient to know their taxonomic position. Thus it will be possible to derive trait characteristics for taxa that are not documented in the trait database, based on their taxonomic relationships.

## Trait characteristics 

The following species traits are used to estimate the fishing impact on species density:

-   The *age* at maturity of species is used to estimate their "intrinsic rate of natural increase" ($r_i$).
-   The *depth* at which species live determines their vulnerability to bottom trawling: it is used to derive the model's "depletion parameter" ($d_i$).

## Fishing intensity

The model also needs the *fishing intensity* for a certain area, for instance expressed as the annual swept area ratio (SAR); this is the cumulative area contacted by a fishing gear over one year and per surface area.

From the swept area ratio (hereafter denoted as $S$, units $yr^{-1}$), we calculate the time in between fishing events ($yr$) as $\frac{1}{S}$.

# Data availability 

## Community composition

For the subsequent analyses, we use macrofauna density data that have been compiled in the frame of the EMODnet biology project, (<https://www.emodnet-biology.eu>).

The monitoring data from the Dutch part of the North sea (called the MWTL monitoring data) comprise both species densities and biomass for `r nrow(MWTL$stations)` stations, that were sampled during `r length(unique(MWTL$density$year))` years, extending from `r min(MWTL$density$year)` till `r max(MWTL$density$year)`.

Sampling was performed yearly at first, then less frequently.
Calculations in this document are based on yearly averages.

The MWTL data are available from the R-package *Btrait*, and are in a list called *MWTL* (within R, type: *?MWTL* to open the help file for these data). 

The composition of the MWTL data is in Appendix 1.

```{r, echo=FALSE}
knitr::kable(data.frame(metadata(MWTL)), 
             caption = "TABLE: Metadata of the MWTL dataset",
             row.names = FALSE)
```

```{r, echo=FALSE}
knitr::kable(head(MWTL$density[, 1:6], n =3), 
             align="l", caption = "TABLE: MWTL$density (first 3 rows)",
             row.names = FALSE)
```

## Taxonomic data

There are `r length(unique(MWTL$density$taxon))` different taxa in the MWTL data.

The taxonomic information for all taxa, has been derived from the WoRMS database and is available from package *Btrait*, in dataset *Taxonomy*: 

```{r, echo=FALSE}
knitr::kable(head(Taxonomy, n =3), align="l", 
             caption = "TABLE: Dataset Taxonomy (first 3 rows)",
             row.names = FALSE)
```

## Trait data

We use the traits compiled by NIOZ (called *Traits_nioz*) for age at maturity, and living depth.
These trait data (mainly on species level) were compiled by Beauchard et al. (2021) and Beauchard et al. (2023).
This database records `r length(unique(metadata(Traits_nioz)$trait))` different traits from `r nrow(Traits_nioz)` taxa.

All traits are fuzzy-coded. The data structure is in appendix 2. 

```{r, echo=FALSE}
knitr::kable(head(Traits_nioz, n = c(3,5)), 
             caption="TABLE: First part of dataset Traits_nioz", 
             digits=2, align = "l", row.names = FALSE)
```

```{r, echo = FALSE}
# merge coordinates with abiotic information
MWTLstats      <- merge(MWTL$station, MWTL$types, 
                        by = "station")  # for area
MWTLstats      <- merge(MWTLstats, MWTL$abiotics, 
                        by = "station")  # for sar
MWTLstats$area <- factor(MWTLstats$area)
```

The figure below shows the position and area of the MWTL stations, and the swept area ratios (sar) that were derived from OSPAR (ICES 2018).

```{r, fig.width=10, fig.height=4.5, fig.cap="FIG: MWTL sampling stations with indication of the main areas, fishing intensity and gear penetration depth. '+' denote the two stations that will be showcased.", echo = FALSE}
par(mfrow = c(1, 3), las = 1)

# add coordinates to abiotic information
MWTLstats      <- merge(MWTL$station, MWTL$types, 
                        by = "station")  # for area
MWTLstats      <- merge(MWTLstats, MWTL$abiotics, 
                        by = "station")  # for sar
MWTLstats$area <- factor(MWTLstats$area)

with(MWTLstats, 
   map_legend(x = x, y = y, colvar = as.integer(area),
             contours = MWTL$contours, draw.levels = TRUE, col.levels = "grey",
             main = "Area", 
             legend = list(x      = "bottomright", side = 0, 
                           legend = levels(MWTLstats$area), 
                           col = 2:5, pch = 15),
             pch = 15, scale = "none", cex = 2, col =  2:5) 
    )
selected <- subset(MWTLstats, subset  = station %in% c("BREEVTN02", "ROTTMPT3"))
points(selected$x, selected$y, pch = "+")
legend("bottomleft", legend = c("BREEVTN02", "ROTTMPT3"), pch = "+", 
       fill = c(2, 5), col = "black", pt.cex = 1, cex = 0.7)

with(MWTLstats, 
   map_MWTL(x = x, y = y, colvar = sar, 
            main = "Fishing effort", clab = "sar, /yr", 
            log = "c", col = Colramp, cex = 2, pch = 15, 
            type = "legend")
)

with(MWTLstats, 
   map_MWTL(x = x, y = y, colvar = gpd, 
            main = "Gear penetration depth", clab = "gpd, cm", 
            col = Colramp, cex = 2, pch = 15,
            type = "legend")
)
```

\newpage

# Model steps

The model is applied in several steps:

* the taxon densities per station are prepared, and a list of taxa in the data is generated,
* the traits recording age at maturity and the depth distribution for each taxon are extracted,
* the fishing parameters for each station are extracted
* the parameter values are estimated based on the traits, fishing parameters, and taxon density 
* the model is run 

# The MWTL density

The MWTL data are gathered over several years, and are in data.frame *MWTL$density*. 
We calculate the density for all stations (descriptor) and taxa, averaged over all years, using *Btrait* function *get_density*.

```{r}
MeanMWTL <- 
  get_density(data        = MWTL$density,
              descriptor  = station,  
              taxon       = taxon,
              value       = density,
              averageOver = year)
```

The taxa whose traits need to be found are extracted:

```{r}
MWTLtaxa <- sort(unique(MeanMWTL$taxon))  
```

# Extracting taxon traits

Taxon traits that we need are in the nioz trait database (*Traits_nioz*), from R-package *Btrait*. 
From this database, we use:

-   the depth distribution (fractional occurrence in each depth layer),
-   the age at maturity of a species

## Substrate depth distribution 

Depth distribution is a fuzzy-coded trait in the trait database.
For use in the model we need to keep it fuzzy coded.

The meaning of the nioz traits is stored in its metadata.
We look for the trait names that relate to depth distribution:

```{r}
meta   <- metadata(Traits_nioz)  # description of trait modalities

meta.D <- subset(meta,           # names of trait modalities for substratum depth
                 subset = (trait == "Substratum depth distribution"))
```

```{r, echo = FALSE}
knitr::kable(meta.D, 
             caption = "TABLE: Depth distribution traits in the NIOZ trait database",
             align = "l", row.names = FALSE)
```

We then extract the depth traits for each taxon in the MWTL density data set.
As not all species in the MWTL dataset have traits assigned, we use the taxonomic information (in dataset *Taxonomy*) to estimate the traits also for unrecorded species.

```{r}
traits.D <- get_trait(
                taxon    = MWTLtaxa, 
                trait    = Traits_nioz[, c("taxon", meta.D$colname)], 
                taxonomy = Taxonomy)

colnames(traits.D)[-1] <- meta.D$modality     # suitable column names
```

```{r, echo = FALSE}
knitr::kable(head(traits.D, n = 3), 
     caption = "TABLE: Depth distribution of species in the MWTL dataset, (first part)",
             row.names = FALSE)
```

Of the `r length(MWTLtaxa)` taxa in the MWTL dataset, only `r sum(MWTLtaxa %in% Traits_nioz$taxon)` taxa are documented in the trait database. The traits for the remaining taxa were estimated based on taxonomic closedness.
Of these taxa, only `r length(attributes(traits.D)$notrait)` could not be assigned a trait.
These were: `r paste(attributes(traits.D)$notrait, collapse = ", ")`.
As these taxa are not true macrofauna, they can be removed from the dataset.

```{r}
Notrait  <- attributes(traits.D)$notrait
MWTLtaxa <- MWTLtaxa [! MWTLtaxa %in% Notrait]
MeanMWTL <- MeanMWTL [! MeanMWTL$taxon %in% Notrait, ]
```

## Age at maturity 

We perform a similar procedure to extract the age at maturity for each taxon in the density data. 
The age at maturity is estimated as the *average* of the fuzzy values.

We first find the relevant trait names in the trait database:

```{r}
meta.A <- subset(meta,     
                 subset = (trait == "Age at maturity"))
```

```{r, echo = FALSE}
knitr::kable(meta.A, 
             caption = "TABLE: Age at maturity traits in the NIOZ trait database",
             row.names = FALSE)
```

We then calculate the average age at maturity. To calculate the average, we pass the *trait_class* and *trait_score* while extracting the traits.

```{r}
traits.A <- get_trait( 
               taxon       = MWTLtaxa, 
               trait       = Traits_nioz[, c("taxon", meta.A$colname)], 
               trait_class = meta.A$trait, 
               trait_score = meta.A$value, 
               taxonomy    = Taxonomy)
```

```{r, echo = FALSE}
knitr::kable(head(traits.A, n = 3), 
    caption = "TABLE: Mean age at maturity of species ([yr]) in the MWTL dataset (first part)",
    digits = 2, align = c("l", "c"), row.names = FALSE)
```

# Merged trait-density data.frame

We merge these two trait datasets in one data.frame, merging by the taxon.

```{r}
TraitsAll <- merge(x  = traits.A, 
                   y  = traits.D, 
                   by = "taxon")
```

and then combine it with the species density data

```{r}
MWTLtrait <- merge(x  = MeanMWTL, 
                   y  = TraitsAll,
                   by = "taxon")

```

# Fishing parameters

Each station has a specific fishing intensity (swept area ratio, *sar*), and mean gear penetration depth (*gpd*). This information is kept in data.frame MWTL$abiotics (from R-package Btrait).

The MWTL data.frame are merged with the swept area ratio, and with the gear penetration depth, merging on column *station*.

```{r}
MWTLtrait <- merge(x  = MWTLtrait, 
                   y  = subset(MWTL$abiotics, 
                               select = c(station, sar, gpd)),
                   by = "station")
```

# Parameter values

## Intrinsic rate of natural increase

The intrinsic rate of natural increase is specific for a species, and can be estimated using function *par_r*. This parameter is added to the MWTL data.frame.

```{r}
MWTLtrait$r <- par_r(age_at_maturity = MWTLtrait$Age.at.maturity)
```

## Depletion fraction d

The depletion rate depends on the depth occurrence of the species (*fDepth*), and on the gear penetration depth (*gpd*). 

The latter differs among stations, so the species depletion fraction will vary with the station.

We also need to input the upper position of each depth layer (*uDepth*).

```{r}
MWTLtrait$d <- par_d(
  gpd    = MWTLtrait$gpd,
  
  fDepth = MWTLtrait[, c( "0", "0-5", "5-15", "15-30", ">30")],
  uDepth =             c(   0,     0,     5,      15,    30)
)
```

## Carrying capacity K

Te carrying capacity is estimated from the model itself. 
We pass the species *density*, the fishing intensity (*sar*), the rate of increase (*r*), and the depletion fraction (*d*):

```{r}
MWTLtrait$K <- par_K(
  
  density = MWTLtrait$density,
  sar     = MWTLtrait$sar,
  
  r       = MWTLtrait$r,
  d       = MWTLtrait$d
)
```

The data that we will use for fishing impact assessment is:

```{r, echo=FALSE}
knitr::kable(head(MWTLtrait[, c("station", "taxon", "r", "d", "K", "sar", "gpd")], n = 3), 
            caption="TABLE: First part of data used for fishing assessment", 
            digits=2, align = "l", row.names = FALSE)
```

## Species for which the model does not find a solution

These are species where the depletion is too high and/or the intrinsic rate of increase is too low, so that it cannot persist in the environment under the current fishing pressure. This causes an anomalously high carrying capacity compared to observed density. Such occurrences are removed from the data set.

```{r}
MWTLtrait$DK <- MWTLtrait$density/MWTLtrait$K
ii <- which(is.na(MWTLtrait$DK)) 
NAS <- subset(MWTLtrait, subset = is.na(DK))

MWTLtrait <- MWTLtrait[-ii,]
```

There are `r nrow(NAS)` such species, the highest observed density being`r max(NAS$density)`. These taxa are removed from the analysis. 

```{r}
TT <- data.frame(table(NAS$taxon))
TM <- aggregate(NAS$density, list(NAS$taxon), FUN=max)
TT <- merge(TT, TM, by = 1)
names(TT) <- c("taxon", "occurrences", "max density")
knitr::kable(TT, caption = "Taxa for which the model could not be solved")
``` 
\newpage

# Running the fishing model

We run the model for `r nrow(MWTLtrait)` species-station occurrences. The model is run for 30 years, of which 20 years of fishing (and thus 10 years where fishing stopped). 

We provide taxon names to label the output columns

```{r}
times <- seq(from = 0, to = 30, length.out = 3000)

system.time(
dens  <- density_perturb(parms        = MWTLtrait, 
                         times        = times, 
                         taxon_names  = paste(MWTLtrait$station, 
                                              MWTLtrait$taxon, sep = "_"),
                         tend_perturb = 20)
)
```

The result is a matrix with `r nrow(dens)` rows and `r ncol(dens)` columns, the first of which contains the time, followed by the density trajectories of all species, and for all stations. 

```{r, echo=FALSE}
knitr::kable(head(dens, n = c(3, 4)), 
            caption="TABLE: First part of model output", 
            digits=2, align = "l", row.names = FALSE)
```

## Summaries

The total density per station is calculated as rowSums per station.

```{r}
Dens  <- data.frame(dens[, -1])  # remove first column
Total <- sapply(split.default(Dens, MWTLtrait$station), # split in stations
                FUN = rowSums, na.rm = TRUE)            # take row sums
```

Then relative values are calculated by dividing all columns (MARGIN = 2) by their maximum (STATS).

```{r, fig.width=8, fig.height=11}  
Drel <- sweep(x      = Total, 
              MARGIN = 2, 
              STATS  = apply(Total,            # max of each column
                             MARGIN = 2, 
                             FUN    = max), 
              FUN    = "/")                    # divide by the max
```

## Some results

We show the results for the 20 most abundant species in two stations.

```{r, fig.width=9, fig.height=9, fig.cap="FIG: Density trajectories of 5 most abundant species from two stations", echo = FALSE}
nsel <- 5
BRtrait   <- subset(MWTLtrait, 
                    subset = station == "BREEVTN02")
B02        <- which(MWTLtrait$station == "BREEVTN02")
BREEVTN02  <- dens[, c(1, 1+B02)]
colnames(BREEVTN02)[-1] <- BRtrait$taxon

BRtrait   <- BRtrait[order(BRtrait$K, decreasing = TRUE), ]
BREEVTN02 <- subset(BREEVTN02, select = c("times", BRtrait$taxon[1:nsel]))


BRtrait   <- subset(MWTLtrait, 
                    subset = station == "ROTTMPT3")
R03       <- which(MWTLtrait$station == "ROTTMPT3")
ROTTMPT3  <- dens[, c(1, 1 + R03)]
colnames(ROTTMPT3)[-1] <- BRtrait$taxon

BRtrait   <- BRtrait[order(BRtrait$K, decreasing = TRUE), ]
ROTTMPT3 <- subset(ROTTMPT3, select = c("times", BRtrait$taxon[1:nsel]))

par(mfrow = c(2, 1), las = 1)

matplot(x    = BREEVTN02[, 1], y = BREEVTN02[, -1], 
        main = "Station BREEVTN02", 
        xlab = "year", ylab = "density, ind/m2",
        type = "l", log = "", lty = 1, lwd = 2, col = 1:nsel)
legend("right", col = 1:5, lty = 1, lwd = 2, cex = 0.75,
       legend = colnames(BREEVTN02)[-1])
matplot(x    = ROTTMPT3 [, 1], y = ROTTMPT3 [, -1], 
        main = "Station ROTTMPT3", 
        xlab = "year", ylab = "density, ind/m2",
        type = "l", log = "", lty = 1, lwd = 2, col = 1:nsel)
legend("right", col = 1:5, lty = 1, lwd = 2, cex = 0.75,
       legend = colnames(ROTTMPT3)[-1])
```

The average density reduction due to fishing in each of the stations is estimated:

```{r, fig.width=5, fig.height=7, fig.cap="FIG: Average fishing impacts for the MWTL stations", echo = TRUE}
# fishing reduction for all species
DDrel <- apply(X      = dens[, -1], 
               MARGIN = 2, 
               FUN    = function(x) 1-min(x)/max(x))

# averaged per station
DDstat <- aggregate(x   = list(p = DDrel), 
                    by  = list(station = MWTLtrait$station), 
                    FUN = mean)

MWTLstat <- merge(MWTL$stations, DDstat,   # add station coordinates 
                  by = "station")

with(MWTLstat, 
    map_MWTL(x    = x, y = y, colvar = p, 
             main = "Fractional reduction ", clab = "-", 
             col  = Colramp, cex = 2, pch = 15, log = "",
             type = "legend")
   )
```
\newpage


# Appendix

## Structure of the MWTL data

The MWTL list contains several data.frames. We will use mainly the data.frames called *density*, *stations*, and *abiotics* (for fishing intensities). 
The structure of these data is:

```{r, echo=FALSE}
knitr::kable(metadata(MWTL$stations), 
             caption = "TABLE: MWTL$station", digits=2, align = "l",
             row.names = FALSE)
knitr::kable(metadata(MWTL$density), 
             caption = "TABLE: MWTL$density", digits=2, align="l",
             row.names = FALSE)
knitr::kable(tail(metadata(MWTL$abiotics), n=7), 
             caption = "TABLE: MWTL$abiotics (related to fishing)", align="l",
             row.names = FALSE)
```

## Structure of the Trait data

```{r, echo=FALSE}
knitr::kable(head(metadata(Traits_nioz)), 
             caption="TABLE: Metadata of Traits_nioz", digits=2, align = "l",
             row.names = FALSE)
```



\newpage

# References

Beauchard O, Brind'Amour A, Schratzberger M, Laffargue P, Hintzen NT, Somerfield PJ, Piet G (2021) A generic approach to develop a trait-based indicator of trawling-induced disturbance.
Mar Ecol Prog Ser 675:35-52.
<https://doi.org/10.3354/meps13840>

Beauchard, O., Murray S.A. Thompson, Kari Elsa Ellingsen, Gerjan Piet, Pascal Laffargue, Karline Soetaert, 2023.
Assessing sea floor functional diversity and vulnerability.
Marine Ecology Progress Series v708, p21-43, <https://www.int-res.com/abstracts/meps/v708/p21-43>

ICES Technical Service, Greater North Sea and Celtic Seas Ecoregions, 29 August 2018 sr.2018.14 Version 2: 22 January 2019 <https://doi.org/10.17895/ices.pub.4508> OSPAR request on the production of spatial data layers of fishing intensity/pressure.

R Core Team (2024). R: A language and environment for statistical computing.
R Foundation for Statistical Computing, Vienna, Austria.
URL <https://www.R-project.org/>.

Soetaert Karline, Olivier Beauchard (2024).
Btrait: Working with Biological density, taxonomy, and trait composition data.
R package version 1.0.1.

Soetaert Karline, Olivier Beauchard (2024).
Bfiat: Bottom Fishing Impact Assessment Tool.
R package version 1.0.1.

Verhulst, Pierre-François (1845).
"Recherches mathématiques sur la loi d'accroissement de la population" [Mathematical Researches into the Law of Population Growth Increase].
Nouveaux Mémoires de l'Académie Royale des Sciences et Belles-Lettres de Bruxelles.
18: 8.


